|

### 2021-11-23 05-Chp-CPRD-Analysis.Rmd
End of day:

<!-- Include the early analysis we did here, that did not work as a result of limited scope. Could include some of the genetic work also. Will additionally need to put the relevant materials from the paper in the  -->

@@

<!-- ## Additional ideas -->

<!-- One of the problems I'll point out with the trials in the systematic review chapter (Chapter \@ref(sys-rev-heading)) is that they only included people with a high cardiovascular risk - we are kind of doing the same by using elevated test cholesterol results as the index event. -->

<!-- Might be able to get around this by pointing out that you need some index event to be able to do this analysis (?) -->

<!-- __Things we tried__ -->

<!-- - Conditioning entry to the cohort on QRISK score (as defined by codes) rather than by . Need to be able to demonstrate here that not many meeting our original entry criteria actually go on to have a statin quickly. Need total number of those starting and average time to start. -->
<!-- - Positive controls. Need to be able to explain why these controls have such wildly increased results. Cross-reference these results with the Hernan paper. -->
<!-- - Competing risk analysis, with death as a competing risk. Need numbers of deaths in each group, and preliminary results -->
<!-- - Allowing for time-varying confounders for binary covariates. -->

<!-- __Things we could try__ -->

<!-- - Work through data using positive control to see if the way it is being added is what is causing the issues. -->
<!-- - Examine Cramer et al. thesis and see what they did -->
<!-- - Marginal structural models approach (a la J. Sterne) -->

@@

<!-- TODO Check all fig/table cross-references evaluate -->

@@

Electronic health record (EHR) databases are large collections of medical data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standardised coding system. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the length of time for which participants are followed. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

@@

Little evidence for an effect of lipid-regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of lipi-regulatin agent called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia was associated with lipid-regulating agent use.

@@

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia as identified by the systematic review. Specifically, this involved a concerted effort to address immortal time bias though use of a Cox proportional hazards analysis, incorporating a time-varying treatment indicator.[@suissa2008] This approach provides a evidence source at risk of a distinct bias due to the alternative analytical strategy. The results from this analysis will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

<!-- TODO, ensure code list position is accurately documented in  -->

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data. Records pre-dating the 1995 cut-off were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability are thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: (a) an outcome of interest; (b) death; (c) end of follow-up (29 February 2016); (d) last registration date with their GP practice; or (e) the last CPRD collection date for their practice. Participants were ineligble for our cohort if they were less than 40 years of age (as these patients are are less likely to be prescribed a LRA), had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

@@

<!----------------------------------------------------------------------------->
(ref:covariateDef-caption) Definition of covariates adjusted for in the full-adjusted model. The codelists used to define the majority of these covariates were originally created for use in a previously published analysis.[@walker2020] while others were built on or adapted from previous published work,[@khan2010; @taylor2016; @wright2017].

@@

Demographic covariates adjusted for included age and gender. Age was calculated at date of entry into the cohort, and adjusted for via its use as the time axis for the Cox model (see Section \@ref(cprd-time-axis)). Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

@@

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] Inclusion of this index allowed me to attempt to adjust for the general health of patients included in the analysis.

Codelists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

@@

Immortal time bias describes two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010] Following the example displayed in the figure, the unexposed group are followed from the date of a diagnosis, while the exposed group is followed from date of a prescription. In this scenario, the time between diangosis and prescription for the exposure group is missing, and any events that occur in the exposed group prior to the prescription will be inappropriately excluded from the analysis.

@@

(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panal B).

@@

This analysis attempted to address this issue by following all participants from a common index date (defined as earliest of: (a) date of raised cholesterol test results; (b) hypercholesterolemeia diagnosis; or (c) LRA prescription). Following a recommended approach to addressing the second form of immortal time bias, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

@@

As detailed in the introduction, the properties of statins may be important in their effect based on the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(intro-statins)).[@sierra2011] To investigate this, I further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins.

@@

#### Comparing dementia codelists

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a published study by Smeeth _et al_.[@smeeth2009]

@@

The results of three models adjusted for age only, age and sex, and the full covariates respectively, are presented in Figure \@ref(fig:unadjustedComparisonFig). These models were used to estimate the impact of adjustment for additional covariates. Note that obtaining an completely unadjusted model is not possible, as age was used in the Cox model as the time scale.

@@

<!----------------------------------------------------------------------------->
(ref:unadjustedComparisonFig-cap) Results of three models adjusting for a different set of covariates.

@@

On the assumption that this variation could be caused by changes in the frequency of codes used to define probable AD in the cohort, I performed a _post-hoc_ investigation of the frequency of each diagnoses stratified by year of entry (Table \@ref(tab:diagnosisType-table)). While the frequency of outcomes declines in more recent strata, likely due to the limited follow-up inherent to these groups, this decline in frequency is relatively constant across the dementia subtypes.

@@

In the cohort, statins with lipophilic properties were much more frequently prescribed than hydrophilic statins (Table \@ref(tab:statinTypeTable-table)). Additionally, there is evidence for a increasing tendency to favour hydrophilic statins in recent years, as the proportion of lipophilic statins prescribed fell from 18.2% in 1996-2000 to <1% in 2011-2016.

@@

#### Impact of dementia code lists {#comparing-codelists}

@@

When using the Smeeth _et al._ code lists to define dementia outcomes, effect estimates of `r smeeth_azd_text` and `r smeeth_oth_text` were obtained for the Alzheimer's disease and non-Alzheimer's ("other") dementia outcomes, respectively.

@@

<!-- TODO Could include part here on the propensity score analysis trimming, and the fact that trimmming using the co-variates that we had access to  -->

<!-- TODO Can also discuss the fact that off-target effects are much easier to get an accurate answer from as they are not as suceptible to confounding by indication -->

@@

<!-- QUESTION To what extent should I cross-reference with the other literature here? -->

@@

<!-- TODO Need to discuss fibrate results as part of qualitative triangulation -->

@@

<!-- IDEA Could include some text here about how you are more likely to get an accurate result when looking at off-target effects -->

@@

__[Feedback requested: How much should I be cross-referencing with the systematic review at this stage? Or should I keep it self-contained, and then provide a comparison across the review/CPRD/IPD analyses as the first part of the triangulation chapter?]__

@@

<!-- IDEA Cite study/paper mentioned by GDS about getting better estimate for effect when dealing with off-target outcomes -->

A likely explanation for the observed increased risk of dementia outcomes with a vascular component (e.g. vascular and other dementias) with lipid-regulating agent use, and an important limitation of this analysis, is residual "confounding by indication". While the term has been used to describe different source of bias in epidemiological analyses,[@salas1999] it is used here to described the role of risk factors that both prompt treatment and increase the risk of the outcome, thus causing a distorted positive association between the treatment and outcome (see Figure \@ref(fig:indicationBias)). In causal inference nomenclature, statins and dementia are said to be _d_-connected as there is an open "backdoor" path between them via the uncontrolled confounders.[@suttorp2015] In the context of this analysis, the confounding variable (or, more likely, variables) would prompt prescription of statins but also represent a vascular risk factor that contributes to the development of the vascular dementia.

@@

Conditioning entry into the study on being either “at-risk” or already diagnosed with hypercholesterolemia was employed in a pre-emptive attempt to mitigate confounding by indication, but evidence from the control outcomes suggests this was unsuccessful. The slight harmful effect for the backpain outcome is substantially smaller that that observed for the ischemic heart disease outcome, indicating that the majority of the uncontrolled confounding is likely related to vascular factors.

@@

Given the absence of vascular dementia in the published literature, as highlighted in the previous chapter (see Section \@ref(sys-rev-pub-bias)), the unexpected increase in vascular dementia risk with statin use is particularly interesting  It is possible that previous research encountered similar methodological issues to this analysis, and via a publication bias mechanism where unexpected or assumedly incorrect results are less likely to be published, the results from these analyses did not make it into the evidence base.

<!-- #### Time period effect in probable Alzheimer's disease -->

<!-- An expected explanation for the time period effect observed for the Alzheimer's disease outcome, where statins appear more protective in participants who entered the cohort at a later date, was an change in how different dementia sub-types were diagnosed over time. For example, if the observed trend was due to increasing differential misclassification with vascular dementia, then we would expect vascular dementia to appear increasingly harmful in more recent time periods which is not the case. However, when investigated via tabulation of dementia outcomes by grouped year of entry (Table \@ref(tab:diagnosisType-table)), no corresponding inverse trend is observed for any of the other dementia outcomes.  -->

<!-- Similarly, the absence of  -->

@@

An initial interpretation of the the different association observed in the two groups was that the lipophilic statins may be more potent, and so are prescribed to patients with a higher underlying vascular load, leading to increased confounding by indication in this group. However, the statin with the strongest lipid lowering effect that is available via the NHS, rosuvastatin, is hydrophobic.

__[Yoav, are there are particular indications/contraindications for statins with different properties? And any general comments on the interpretation of these findings would be welcome!]__

@@

TThis finding disagrees with the results of the original analysis, which found evidence for a protective effect of statin use on all-cause dementia (`r estimate(0.81, 0.69,0.96,type = "HR")`) and non-AD  dementia (`r estimate(0.82,0.69, 0.97,type = "HR")`), but little evidence of an effect on AD (`r estimate(0.81,0.49, 1.35, type="HR")`).

@@

The primary strength of this analysis is the relative size of the CPRD and length of follow-up. Having reviewed the existing literature, as identified by the systematic review in Chapter \@ref(sys-rev-heading), this analysis of `r comma(attrition[14,1])` participants is one of the largest available studies of this research question. Additionally, this analysis followed LRA users and non-users from a common index date, using a time-updating treatment indicator to correctly assign time-at-risk to the exposed and unexposed groups. This approach has been less commonly used in the literature and allows for the mitigation of potential immortal time bias. Finally, it is one of the few that provide evidence on the association of lipid-regulating agent use and vascular and other dementias.

However, the findings of this analysis are subject to several limitations in addition to the confounding by indication discussed above. There is a strong possibility of differential misclassification of dementia-related conditions based on the exposure.[@porta2014] As an illustrative example, those with memory complaints may be more likely to be classified as vascular dementia than Alzheimer's disease if their medical records contains prescriptions for lipid-regulating agents. Further, there is a potential for general non-differential misclassification of the outcome due to the varying positive predictive value of electronic health record code lists to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]
<!-- though there is some encouraging evidence that a dedicated coding exercise increases the identification of dementia case in primary care database and does not add a substantial administrative burden.[@russell2013] -->

Misclassification of outcomes is not the only issue introduced by the use of EHR codes to define outcomes. Comparing and contrasting between different studies is particularly difficult because of the impact that the use of different code list can have on the analysis, as illustrated by the discrepancy between the results when using the codes lists defined for this study and those used by Smeeth _et al_. This presents a particular challenge in comparing research across different time-periods and coding systems, particularly the unexpected results I obtained for the vascular and other dementias outcomes.

A further limitation stems from the possibility of uncontrolled confounding due to genetic factors. Number of ApoE $\epsilon4$ alleles represents the strongest genetic risk factor for Alzheimer's disease, but also substantially increases LDL-c levels,[@bennet2007] potentially prompting treatment with a statin or other lipid regulating agent. I was unable to control for ApoE genotype in this analysis as I did not have access to genetic data on participants. As a result, any protective association between LRA use and the Alzheimer's disease outcomes may be masked by residual negative confounding by ApoE.

@@

### Conclusions

This chapter has provided new evidence on the association of lipid-regulating agents with incidence of all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementia. It made use of a large scale electronic health record database, the CPRD, and employed a time-varying Cox proportional hazards model to account for potential immortal time bias.

It found little evidence for an effect of lipid-regulating agents on probable or possible Alzheimer’s disease. However, lipid-regulating agent use was associated with an increased risk of all-cause, vascular and other dementias. In all cases, the estimated associations were driven by those observed in the statin subgroup, which comprised the majority of exposed participants in this cohort.

This chapter attempted to account for important sources of bias and provide a comparison with other available literature, as identified in the systematic review presented in Chapter \@ref(sys-rev-heading). However, there is a strong potential for uncontrolled confounding by indication and differential misclassification of the outcome on the basis of exposure, which raises questions about the findings, in particular the unexpected increase in risk of vascular dementia associated with statin use. The presence of bias is supported by our findings for the negative and positive control outcomes used, which provide some evidence of uncontrolled vascular confounders that may both prompt LRA prescription and increase risk of dementia outcomes with a vscular component. Future research using large scale electronic health records should aim to address these limitations, potentially by using a analytical design that more closely emulates a trial.[@danaei2013]

Regardless, this analysis has provided an additional source of evidence for the triangulation exercise presented in Chapter \@ref(discussion-heading). In the following Chapter, the dataset described here is incorporated along with several other datasets as part of an individual patient data analysis to investigate the association of blood lipid levels with incidence of dementia outcomes directly, rather than via the proxy of lipid-regulating treatment.

>>>>

@@

@@

@@

Electronic health record (EHR) databases are large collections of medical data, used to manage patient administration and care. Under these systems, whenever a patient attends their GP, their clinical data is recorded in a central database using a standardised coding system. These databases have several advantages over traditional methods of data collection, including the number of people they contain and the relatively low cost of data collection via routine care. This is particularly important when studying diseases such as dementia, which may begin to develop in patients long before symptoms are seen.

@@

Little evidence for an effect of lipid-regulating agents effect on the risk of Alzheimer's disease was found, with the exception of a slightly increased risk in those prescribed a certain type of lipid-regulating agent called fibrates. In contrast, I found an increased risk of vascular and other (i.e. non-Alzheimer's) dementia was associated with lipid-regulating agent use.

@@

Secondly, and in a similar vein, the analysis intentionally takes a different analytical approach to that most commonly used to examine the effect of statins on dementia as identified by the systematic review. Specifically, this involved a concerted effort to address immortal time bias though use of a Cox proportional hazards analysis, incorporating a time-varying treatment indicator.[@suissa2008] By employing this approach, the analysis provides an evidence source at risk of a distinct bias. The results from this analysis will be incorporated into the triangulation exercise presented in Chapter \@ref(discussion-heading).

@@

<!-- TODO ensure code list position is accurately documented in Appendix -->

@@

This analysis included all participants registered at a participating practice between 1 January 1995 and 29 February 2016 who had a flag for "research quality" data (as defined by the CPRD). Records pre-dating the 1995 cut-off were included in the original CPRD extract obtained for this analysis. However, these were excluded from the analysis as data quality and reliability are thought to be higher after this date.[@wolf2019] Additionally, individuals with less than 12 months of continuous records prior to cohort entry were excluded, making the effective start date of the cohort 1 January 1996.

@@

The index date for a participant was defined as the date where the first relevant code or test result was recorded on their clinical record, and participants were followed up until the earliest of: (a) an outcome of interest; (b) death; (c) end of follow-up (29 February 2016); (d) last registration date with their GP practice; or (e) the last CPRD collection date for their practice. Participants were ineligible for our cohort if they were less than 40 years of age (as these patients are  less likely to be prescribed a LRA), had less than 12 months of "research quality" data, were simultaneously prescribed more than one lipid-regulating agent (due to the difficult of assigning these to a single exposure group), or were diagnosed with an outcome of interest before or on the date of the index event (i.e. had less than one full day of follow-up).

@@

<!-- TODO Need to deal with the fact that the ordering is not accounted for, which could be critised -->

@@

<!----------------------------------------------------------------------------->
(ref:covariateDef-caption) Definition of covariates adjusted for in the full-adjusted model. The code lists used to define the majority of these covariates were originally created for use in a previously published analysis.[@walker2020] while others were built on or adapted from previous published work,[@khan2010; @taylor2016; @wright2017].

@@

Demographic covariates adjusted for included age and gender. Age in years was calculated at date of entry into the cohort, using the 1st of January of a patients' birth year (the exact date of birth was not provided by CPRD), and was adjusted for via its use as the time axis for the Cox model (see Section \@ref(cprd-time-axis)). Socioeconomic status was proxied using the Index of Multiple Deprivation (IMD) 2010, which draws on seven domains (income; employment; education, skills and training; health and disability; crime; barriers to housing and services; living environment) to create an overall deprivation score for each of 32844 statistical geography areas in England. To help preserve patient privacy, IMD score is only available from the CPRD in twentiles, with 1 indicating the least deprived and 20 indicating the most deprived.  Smoking and alcohol use was determined at index, and participants were categorised as current, former, or never users of each.

@@

Finally, presence of a range of related conditions at baseline were accounted for, including cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. In addition to adjusting for these covariates individually, a Charlson co-morbidity index (CCI) score was calculated for each participant. The CCI is a weighted index that uses presence and severity of a number of conditions to enable adjustment for the general health of a participant in terms of their mortality risk.[@charlson1987new] The conditions considered under the index are: AIDS; cancer; cerebrovascular disease; chronic pulmonary disease; congestive heart disease; dementia; diabetes; diabetes with complications; hemiplegia; metastatic tumour; mild liver disease; moderate liver disease; myocardial infarction; peptic ulcer disease; peripheral vascular disease; renal disease; and rheumatological disease. Inclusion of this index allowed me to attempt to adjust for the general health of patients included in the analysis.

code lists for all covariates can be found in the archived data repository accompanying this analysis (see Section \@ref(cprd-data-avail)).

@@

Immortal time bias describes two distinct but related types of bias. The first presentation, the selection bias aspect (Panel A, Figure \@ref(fig:immortalTimeBias)), occurs when time prior to the exposure is excluded leading to the exposed and control groups being followed up from different time points.[@levesque2010] Following the example displayed in the figure, the unexposed group are followed from the date of a diagnosis, while the exposed group is followed from date of a prescription. In this scenario, the time between diagnosis and prescription for the exposure group is missing, and any events that occur in the exposed group prior to the prescription will be inappropriately excluded from the analysis.

@@

(ref:immortalTimeBias-cap) Diagram illustrating the two presentations of immortal time bias, as a selection bias (Panel A) and a misclassification bias (Panel B).

@@

This analysis attempted to address this issue by following all participants from a common index date (defined as earliest of: (a) date of raised cholesterol test results; (b) hypercholesterolemia diagnosis; or (c) LRA prescription). Following a recommended approach to addressing the second form of immortal time bias, I employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010]

@@

As detailed in the introduction, the properties of statins may be important in their effect based on the ability of lipophilic statins to cross the blood brain barrier (see Section \@ref(intro-statins)).[@sierra2011] As such, I expected that any effects of statins on dementia outcomes would be stronger in the lipophilic as compared to the hydrophilic statin subgroup. To investigate this, I further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins.

@@

#### Impact of using different code lists for defining dementia outcomes

As part of an exploratory analysis of the effect of the choice of code lists on the analysis, I created an alternative Alzheimer's disease and non-Alzheimer's dementia outcome using code lists from a published study by Smeeth _et al_.[@smeeth2009] The intended purpose of this analysis was to assess the robustness of my results to the choice of code list.

@@

The results of three models adjusted for age only, age and sex, and the full covariates respectively, are presented in Figure \@ref(fig:unadjustedComparisonFig). These models were used to estimate the impact of adjustment for additional covariates. Note that obtaining a completely unadjusted model is not possible, as age was used in the Cox model as the time scale.

@@

<!----------------------------------------------------------------------------->
(ref:unadjustedComparisonFig-cap) Results of three models adjusting for a different set of covariates. Note that the x-axis cutoffs (.5,2) are different compared to other plots (0.3,3) to enable greater comparison between the different models.

@@

On the assumption that this variation could be caused by changes in the frequency probable AD diagnoses in the cohort over time, I performed a _post-hoc_ investigation of the frequency of each diagnoses stratified by year of entry (Table \@ref(tab:diagnosisType-table)). While the frequency of outcomes declines in more recent strata, likely due to the limited follow-up inherent to these groups, this decline in frequency is relatively constant across the dementia subtypes.

@@

In the cohort, statins with lipophilic properties were much more frequently prescribed than hydrophilic statins (Table \@ref(tab:statinTypeTable-table)). Additionally, there is evidence for an increasing tendency to favour hydrophilic statins in recent years, as the proportion of lipophilic statins prescribed fell from 18.2% in 1996-2000 to <1% in 2011-2016.

@@

#### Impact of dementia code lists {#comparing-code lists}

@@

When using the Smeeth _et al._ code lists to define dementia outcomes, effect estimates of `r smeeth_azd_text` and `r smeeth_oth_text` were obtained for the Alzheimer's disease and non-Alzheimer's ("other") dementia outcomes, respectively. While direct mapping to the outcomes used in this analysis was not possible, the most comparable are probable Alzheimer's disease (`r probad_text`) and other dementia (`r othdem_text`).

@@

@@

@@

@@

@@

@@

A likely explanation for the observed increased risk of dementia outcomes with a vascular component (e.g. vascular and other dementias) with lipid-regulating agent use, and an important limitation of this analysis, is residual "confounding by indication". While the term has been used to describe different source of bias in epidemiological analyses,[@salas1999] it is used here to described the role of risk factors that both prompt treatment and increase the risk of the outcome, thus causing a distorted positive association between the treatment and outcome (see Figure \@ref(fig:indicationBias)). In causal inference nomenclature, statins and dementia are said to be _d_-connected as there is an open "backdoor" path between them via the uncontrolled confounders.[@suttorp2015] In the context of this analysis, the confounding variable (or, more likely, variables) would prompt both prescription of statins and also represent a vascular risk factor that contributes to the development of the vascular dementia. A similar confounding structure likely exists for ezetimibe, another hypercholesterolemia treatment, providing an explanation for the association of vascular/other dementia but not Alzheimer's disease with this exposure.

@@

Conditioning entry into the study on being either “at-risk” or already diagnosed with hypercholesterolemia was employed in a pre-emptive attempt to mitigate confounding by indication, but evidence from the control outcomes suggests this was unsuccessful. The slight harmful effect for the backpain outcome is substantially smaller than that observed for the ischemic heart disease outcome, indicating that the majority of the uncontrolled confounding is likely related to vascular factors. The slight effect observed for the negative control of backpain could be due to incomplete control for socioeconomic status, as deprivation data was provided in twentiles to preserve patient privacy.[@boruzs2016; @ikeda2019]

@@

Given the absence of vascular dementia in the published literature, as highlighted in the previous chapter (see Section \@ref(sys-rev-pub-bias)), the unexpected increase in vascular dementia risk with statin use is particularly interesting. It is possible that this finding reflects publication bias if previous research encountered similar methodological issues to this analysis, and these results did not make it into the evidence base as unexpected or assumedly incorrect results are less likely to be submitted or accepted for publication.

@@

An initial interpretation of the different associations observed in the two groups was that the lipophilic statins may be more potent, and so are prescribed to patients with a higher underlying vascular load, leading to increased confounding by indication in this group. However, the statin with the strongest lipid lowering effect that is available via the NHS, rosuvastatin, is hydrophobic.

@@

This finding disagrees with the results of the original analysis, which found evidence for a protective effect of statin use on all-cause dementia (`r estimate(0.81, 0.69,0.96,type = "HR")`) and non-AD  dementia (`r estimate(0.82,0.69, 0.97,type = "HR")`), but little evidence of an effect on AD (`r estimate(0.81,0.49, 1.35, type="HR")`).

@@

The primary strength of this analysis is the relative size of the CPRD and length of follow-up. Having reviewed the existing literature, as identified by the systematic review in Chapter \@ref(sys-rev-heading), this analysis of `r comma(attrition[14,1])` participants is one of the largest available studies of this research question. Additionally, this analysis followed LRA users and non-users from a common index date, using a time-updating treatment indicator to correctly assign time-at-risk to the exposed and unexposed groups. This approach has been less commonly used in the literature and allows for the mitigation of potential immortal time bias. Finally, it is one of the few that provide evidence on the association of lipid-regulating agent use and vascular and other dementias, and used negative and positive controls to assess the potential for residual confounding.

However, the findings of this analysis are subject to several limitations in addition to the confounding by indication discussed above. There is a strong possibility of differential misclassification of dementia-related conditions based on the exposure.[@porta2014] As an illustrative example, those with memory complaints may be more likely to be classified as vascular dementia than Alzheimer's disease if their medical records contain prescriptions for lipid-regulating agents. Further, there is a potential for general non-differential misclassification of the outcome due to the varying positive predictive value of electronic health record code lists to identify dementia cases.[@wilkinson2018; @mcguinness2019validity]

Misclassification of outcomes is not the only issue introduced by the use of EHR codes to define outcomes. Comparing and contrasting between different studies is particularly difficult because of the impact that the use of different code lists can have on the analysis, as illustrated by the discrepancy between the results when using the code lists defined for this study and those used by Smeeth _et al_. This presents a particular challenge in comparing research across different time-periods and coding systems, particularly the unexpected results I obtained for the vascular and other dementias outcomes.

A further limitation stems from the possibility of uncontrolled confounding due to genetic factors. The number of ApoE $\epsilon4$ alleles represents the strongest genetic risk factor for Alzheimer's disease, but also substantially increases LDL cholesterol levels,[@bennet2007] potentially prompting treatment with a statin or other lipid regulating agent. I was unable to control for ApoE genotype in this analysis as I did not have access to genetic data on participants. As a result, any protective association between LRA use and the Alzheimer's disease outcomes may be masked by residual negative confounding by ApoE.

@@

### Summary

* In this chapter, I produced new evidence on the association of lipid-regulating agents with incidence of all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementia.

* I found little evidence for an effect of lipid-regulating agents on probable or possible Alzheimer’s disease. However, lipid-regulating agent use was associated with an increased risk of all-cause, vascular and other dementias. In all cases, the estimated associations were driven by those observed in the large statin subgroup.

* I attempted to account for important sources of bias through use of time-varying treatment indicator, though the control outcomes included in the analysis provided evidence for only partially controlled confounding by indication related to vascular factors. Additionally, there was the potential for differential misclassification of dementia subtype on the basis of the exposure. Combined, these biases reduce the confidence in my findings, in particular the unexpected increase in risk of vascular dementia associated with statin use.

* Findings from this analysis are used as an additional source of evidence in the triangulation exercise presented in Chapter \@ref(discussion-heading).
